\subsection{Graphing}\par

{\tmstrong{Solve each system by graphing.}}

\begin{multicols}{4}
  1)\\
	$y = - x + 1$\\
	$y = - 5 x - 3$\par
  ~\par
  5)\\
  $y = - \frac{3}{4} x + 1$\\
	$y = - \frac{3}{4} x + 2$\par
  ~\par
  9)\\
  $y = \frac{5}{3} x + 4$\\
	$y = - \frac{2}{3} x - 3$\par
  ~\par
  13)\\
  $x - y = 4$\\
	$2 x + y = - 1$\par
  ~\par
  17)\\
  $2 x + y = 2$\\
	$x - y = 4$\par
  ~\par
  21)\\
  $9y+6x=36$\\
	$3y-6x=-12$\par
  ~\par
  25)\\
  $3 + y = - x$\\
	$- 4 - 6 x = - y$\par
  ~\par
  29)\\
  $- 12 + x = 4 y$\\
	$12 - 5 x = 4 y$\par
  ~\par
  ~\par
  ~\par

  2)\\
  $y = - \frac{5}{4} x - 2$\\
	$y = - \frac{1}{4} x + 2$\par
  ~\par
  6)\\
  $y = 2 x + 2$\\
	$y = - x - 4$\par
  ~\par
  10)\\
  $y = \frac{1}{2} x + 4$\\
	$y = \frac{1}{2} x + 1$\par
  ~\par
  14)\\
  $6 x + y = - 3$\\
	$x + y = 2$\par
  ~\par
  18)\\
  $x + 2 y = 6$\\
	$5 x - 4 y = 16$\par
  ~\par
  22)\\
  $- 2 y + x = 4$\\
	$2 = - x + \frac{1}{2} y$\par
  ~\par
  26)\\
  $16 = - x - 4 y$\\
	$- 2 x = - 4 - 4 y$\par
  ~\par
  30)\\
  $- 5 x + 1 = - y$\\
	$- y + x = - 3$\par
  ~\par
	~\par
	~\par
	~\par
	
  3)\\
  $y = - 3$\\
	$y = - x - 4$\par
  ~\par
  7)\\
  $y = \frac{1}{3} x + 2$\\
	$y = - \frac{5}{3} x - 4$\par
  ~\par
  11)\\
  $x + 3 y = - 9$\\
	$5 x + 3 y = 3$\par
  ~\par
  15)\\
  $2 x + 3 y = - 6$\\
	$2 x + y = 2$\par
  ~\par
  19)\\
  $2 x + y = - 2$\\
	$x + 3 y = 9$\par
  ~\par
  23)\\
  $2 x - y = - 1$\\
	$3 = - 2 x - y$\par
  ~\par
  27)\\
  $- y + 7 x = 4$\\
	$- y + 7 x = 3$\par
  ~\par
	~\par
	~\par
	~\par
	~\par
	~\par
  ~\par

  4)\\
  $y = - x - 2$\\
	$y = \frac{2}{3} x + 3$\par
  ~\par
  8)\\
  $y = 2 x - 4$\\
	$y = \frac{1}{2} x + 2$\par
  ~\par
  12)\\
  $x + 4 y = - 12$\\
	$2 x + y = 4$\par
  ~\par
  16)\\
  $3 x + 2 y = 2$\\
	$3 x + 2 y = - 6$\par
  ~\par
  20)\\
  $x - y = 3$\\
	$5 x + 2 y = 8$\par
  ~\par
	24)\\
  $- 2 y = - 4 - x$\\
	$- 2 y = - 5 x + 4$\par
  ~\par
  28)\\
  $- 4 + y = x$\\
	$x + 2 = - y$\par
  ~\par
	~\par
	~\par
	~\par
	~\par
	~\par
  ~\par
	
\end{multicols}

\newpage

\subsection{Substitution}\par

{\tmstrong{Solve each system by substitution.}}

\begin{multicols}{4}
  1)\\ 
	$y = - 3 x$\\
	$y = 6 x - 9$\par
  ~\par
  5)\\ 
  $y = 6 x + 4$\\
	$y = - 3 x - 5$\par
  ~\par
  9)\\ 
  $y = 2 x - 3$\\
	$y = - 2 x + 9$\par
  ~\par
  13)\\ 
  $y = - 6$\\
	$3 x - 6 y = 30$\par
  ~\par
  17)\\ 
  $- 2 x + 2 y = 18$\\
  $y = 7 x + 15$\par
  ~\par
  21)\\ 
  $7 x - 2 y = - 7$\\
  $y = 7$\par
  ~\par
  25)\\ 
  $- 2 x - y = - 5$\\
  $x - 8 y = - 23$\par
  ~\par
  29)\\ 
  $3 x + y = 9$\\
	$2 x + 8 y = - 16$\par
  ~\par
  33)\\ 
  $x + 5 y = 15$\\
  $- 3 x + 2 y = 6$\par
  ~\par

  2)\\ 
  $y = x + 5$\\
	$y = - 2 x - 4$\par
  ~\par
  6)\\ 
  $y = 3 x + 13$\\
	$y = - 2 x - 22$\par
  ~\par
  10)\\ 
  $y = 7 x - 24$\\
	$y = - 3 x + 16$\par
  ~\par
  14)\\ 
  $6 x - 4 y = - 8$\\
  $y = - 6 x + 2$\par
  ~\par
  18)\\ 
  $y = x + 4$\\
	$3 x - 4 y = - 19$\par
  ~\par
  22)\\ 
  $x - 2 y = - 13$\\
  $4 x + 2 y = 18$\par
  ~\par
  26)\\ 
  $6 x + 4 y = 16$\\
  $- 2 x + y = - 3$\par
  ~\par
  30)\\ 
  $- 5 x - 5 y = - 20$\\
  $- 2 x + y = 7$\par
  ~\par
  34)\\ 
  $2 x + 3 y = - 10$\\
  $7 x + y = 3$\par
  ~\par
  
  3)\\ 
  $y = - 2 x - 9$\\
	$y = 2 x - 1$\par
  ~\par
  7)\\ 
  $y = 3 x + 2$\\
	$y = - 3 x + 8$\par
  ~\par
  11)\\ 
  $y = 6 x - 6$\\
	$- 3 x - 3 y = - 24$\par
  ~\par
  15)\\ 
  $y = - 5$\\
	$3 x + 4 y = - 17$\par
  ~\par
  19)\\ 
  $y = - 8 x + 19$\\
	$- x + 6 y = 16$\par
  ~\par
  23)\\ 
  $x - 5 y = 7$\\
	$2 x + 7 y = - 20$\par
  ~\par
  27)\\ 
  $- 6 x + y = 20$\\
  $- 3 x - 3 y = - 18$\par
  ~\par
  31)\\ 
  $2 x + y = 2$\\
  $3 x + 7 y = 14$\par
  ~\par
  35)\\ 
  $- 2 x + 4 y = - 16$\\
  $y = - 2$\par
  ~\par
  
  4)\\ 
  $y = - 6 x + 3$\\
	$y = 6 x + 3$\par
  ~\par
  8)\\ 
  $y = - 2 x - 9$\\
	$y = - 5 x - 21$\par
  ~\par
  12)\\ 
  $- x + 3 y = 12$\\
  $y = 6 x + 21$\par
  ~\par
  16)\\ 
  $7 x + 2 y = - 7$\\
  $y = 5 x + 5$\par
  ~\par
  20)\\ 
  $y = - 2 x + 8$\\
  $- 7 x - 6 y = - 8$\par
  ~\par
  24)\\ 
  $3 x - 4 y = 15$\\
  $7 x + y = 4$\par
  ~\par
  28)\\ 
  $7 x + 5 y = - 13$\\
  $x - 4 y = - 16$\par
  ~\par
  32)\\ 
  $2 x + y = - 7$\\
  $5 x + 3 y = - 21$\par
  ~\par
  36)\\ 
  $- 2 x + 2 y = - 22$\\
  $- 5 x - 7 y = - 19$\par
  ~\par
  
\end{multicols}

\begin{multicols}{4}
  37)\\ 
  $- 6 x + 6 y = - 12$\\
  $8 x - 3 y = 16$\par
  ~\par
  
  38)\\ 
  $- 8 x + 2 y = - 6$\\
  $- 2 x + 3 y = 11$\par
  ~\par

  39)\\ 
  $2 x + 3 y = 16$\\
  $- 7 x - y = 20$\par
  ~\par

  40)\\ 
  $- x - 4 y = - 14$\\
  $- 6 x + 8 y = 12$\par
  ~\par	
\end{multicols}


\newpage

\subsection{Addition/Elimination}\par

{\tmstrong{Solve each system by elimination.}}
\label{matrices1}
\begin{multicols}{4}
  1)\\
	$4 x + 2 y = 0$\\
  $- 4 x - 9 y = - 28$\par
  ~\par
  5)\\
	$- 6 x + 9 y = 3$\\
  $6 x - 9 y = - 9$\par
  ~\par 
  9)\\
  $- x - 5 y = 28$\\
  $- x + 4 y = - 17$\par
  ~\par
  13)\\
  $10 x + 6 y = 24$\\
  $- 6 x + y = 4$\par
  ~\par
  17)\\
  $- 7 x + 4 y = - 4$\\
  $10 x - 8 y = - 8$\par
  ~\par
  21)\\
  $- 7 x - 3 y = 12$\\
  $- 6 x - 5 y = 20$\par
  ~\par
  25)\\
  $9 x + 6 y = - 21$\\
  $- 10 x - 9 y = 28$\par
  ~\par
  29)\\
  $- 8 x - 8 y = - 8$\\
  $10 x + 9 y = 1$\par
  ~\par
  33)\\
  $0 = 9 x + 5 y$\\
  $y = \frac{2}{7} x$\par
	~\par
	
  2)\\
  $- 7 x + y = - 10$\\
  $- 9 x - y = - 22$\par
  ~\par
  6)\\
  $5 x - 5 y = - 15$\\
  $5 x - 5 y = - 15$\par
  ~\par
  10)\\
  $- 10 x - 5 y = 0$\\
  $10 x +10 y=30$\par
  ~\par
  14)\\
  $x + 3 y = - 1$\\
  $10 x + 6 y = - 10$\par
  ~\par
  18)\\
  $- 6 x + 4 y = 4$\\
  $- 3 x - y = 26$\par
  ~\par
  22)\\
  $- 5 x + 4 y = 4$\\
  $- 7 x - 10 y = - 10$\par
  ~\par
  26)\\
  $- 4 x - 5 y = 12$\\
  $- 10 x + 6 y = 30$\par
  ~\par
  30)\\
  $- 7 x + 10 y = 13$\\
  $4 x + 9 y = 22$\par
  ~\par
  34)\\
  $- 6 - 42 y = - 12 x$\\
	$x - \frac{7}{2} y = \frac{1}{2} $\par
  ~\par
	
  3)\\
  $- 9 x + 5 y = - 22$\\
  $9 x - 5 y = 13$\par
  ~\par
  7)\\
  $4 x - 6 y = - 10$\\
  $4 x - 6 y = - 14$\par
  ~\par
  11)\\
  $2 x - y = 5$\\
  $5 x + 2 y = - 28$\par
  ~\par
  15)\\
  $2 x + 4 y = 24$\\
  $4 x - 12 y = 8$\par
  ~\par
  19)\\
  $5 x + 10 y = 20$\\
  $- 6 x - 5 y = - 3$\par
  ~\par
  23)\\
  $9 x - 2 y = - 18$\\
  $5 x - 7 y = - 10$\par
  ~\par
  27)\\
  $- 7 x + 5 y = - 8$\\
  $- 3 x - 3 y = 12$\par
  ~\par
  31)\\
  $9 y = 7 - x$\\
  $- 18 y + 4 x = - 26$\par
	~\par
	~\par
	~\par
	~\par
	~\par

  4)\\
  $- x - 2 y = - 7$\\
  $x + 2 y = 7$\par
  ~\par
  8)\\
  $- 3 x + 3 y = - 12$\\
  $- 3 x + 9 y = - 24$\par
  ~\par
  12)\\
  $- 5 x + 6 y = - 17$\\
  $x - 2 y = 5$\par
  ~\par
  16)\\
  $- 6 x + 4 y = 12$\\
  $12 x + 6 y = 18$\par
  ~\par
  20)\\
  $- 9 x - 5 y = - 19$\\
  $3 x - 7 y = - 11$\par
  ~\par
  24)\\
  $3 x + 7 y = - 8$\\
  $4 x + 6 y = - 4$\par
  ~\par
  28)\\
  $8 x + 7 y = - 24$\\
  $6 x + 3 y = - 18$\par
  ~\par
  32)\\
  $21 = - 9 x + 12 y$\\
	$\frac{4}{3} y + \frac{7}{3} x=-1$\par
  ~\par
	~\par
	~\par
	~\par
	~\par
	
\end{multicols}

\pagebreak

\subsection{Three Variables}\par

{\tmstrong{Solve each of the following systems of equation.}}
\label{matrices2}
\begin{multicols}{3}
  1)\\
	$a - 2 b + c = 5$\\
  $2 a + b - c = - 1$\\
  $3 a + 3 b - 2 c = - 4$\par
~\par
  4)\\
  $x + y + z = 2$\\
  $6 x - 4 y + 5 z = 31$\\
  $5 x + 2 y + 2 z = 13$\par
~\par
  7)\\
  $x + y + z = 6$\\
  $2 x - y - z = - 3$\\
  $x - 2 y + 3 z = 6$\par
~\par
  10)\\
  $x + 2 y - z = 4$\\
  $4 x - 3 y + z = 8$\\
  $5 x - y = 12$\par
~\par
  13)\\
  $2 x + y - 3 z = 0$\\
  $x - 4 y + z = 0$\\
  $4 x + 16 y + 4 z = 0$\par
~\par
  16)\\
  $p + q + r = 1$\\
  $p + 2 q + 3 r = 4$\\
  $4 p + 5 q + 6 r = 7$\par
~\par
  19)\\
  $x - y + 2 z = 0$\\
  $x - 2 y + 3 z = - 1$\\
  $2 x - 2 y + z = - 3$\par
~\par

  2)\\
  $2 x + 3 y = z - 1$\\
  $3 x = 8 z - 1$\\
  $5 y + 7 z = - 1$\par
~\par
  5)\\
  $x + 6 y + 3 z = 4$\\
  $2 x + y + 2 z = 3$\\
  $3 x - 2 y + z = 0$\par
~\par
  8)\\
  $x + y - z = 0$\\
  $x + 2 y - 4 z = 0$\\
  $2 x + y + z = 0$\par
~\par
  11)\\
  $- 2 x + y - 3 z = 1$\\
  $x - 4 y + z = 6$\\
  $4 x + 16 y + 4 z = 24$\par
~\par
  14)\\
  $4 x + 12 y + 16 z = 0$\\
  $3 x + 4 y + 5 z = 0$\\
  $x + 8 y + 11 z = 0$\par
~\par
  17)\\
  $x - 2 y + 3 z = 4$\\
  $2 x - y + z = - 1$\\
  $4 x + y + z = 1$\par
~\par
  20)\\
  $4 x - 7 y + 3 z = 1$\\
  $3 x + y - 2 z = 4$\\
  $4 x - 7 y + 3 z = 6$\par
~\par
  
  3)\\
  $3 x + y - z = 11$\\
  $x + 3 y = z + 13$\\
  $x + y - 3 z = 11$\par
~\par
  6)\\
  $x - y + 2 z = - 3$\\
  $x + 2 y + 3 z = 4$\\
  $2 x + y + z = - 3$\par
~\par
  9)\\
  $x + y - z = 0$\\
  $x - y - z = 0$\\
  $x + y + 2 z = 0$\par
~\par
  12)\\
  $4 x + 12 y + 16 z = 4$\\
  $3 x + 4 y + 5 z = 3$\\
  $x + 8 y + 11 z = 1$\par
~\par
  15)\\
  $3 x + 2 y + 2 z = 3$\\
  $x + 2 y - z = 5$\\
  $2 x - 4 y + z = 0$\par
~\par
  18)\\
  $x + 2 y - 3 z = 9$\\
  $2 x - y + 2 z = - 8$\\
  $3 x - y - 4 z = 3$\par
~\par
  21)\\
  $4 x - 3 y + 2 z = 40$\\
  $5 x + 9 y - 7 z = 47$\\
  $9 x + 8 y - 3 z = 97$\par
~\par

\end{multicols}

\begin{multicols}{3}
 \label{matrices3}
  22)\\
  $3 x + y - z = 10$\\
  $8 x - y - 6 z = - 3$\\
  $5 x - 2 y - 5 z = 1$\par
~\par
  25)\\
  $3 x - 4 y + 2 z = 1$\\
  $2 x + 3 y - 3 z = - 1$\\
  $x + 10 y - 8 z = 7$\par
~\par
  28)\\
  $3 x + 2 y = z + 2$\\
  $y = 1 - 2 x$\\
  $3 z = - 2 y$\par
~\par  
  31)\\
  $w + x + y + z = 2$\\
  $w + 2 x + 2 y + 4 z = 1$\\
  $w - x + y + z = 6$\\
	$w - 3 x - y + z = 2$\par
~\par
  
  23)\\
  $3 x + 3 y - 2 z = 13$\\
  $6 x + 2 y - 5 z = 13$\\
  $5 x - 2 y - 5 z = - 1$\par
~\par
  26)\\
  $2 x + y = z$\\
  $4 x + z = 4 y$\\
  $y = x + 1$\par
~\par
  29)\\
  $2 w - 2 x - 2 y + 2 z = 10$\\
	$w + x + y + z = - 5$\\
	$3 w + 2 x + 2 y + 4 z = - 11$\\
	$w + 3 x - 2 y + 2 z = - 6$\par
~\par
  32)\\
  $w + x - y + z = 0$\\
  $- w + 2 x + 2 y + z = 5$\\
	$w - 3 x - y + z = 4$\\
	$2 w - x - y + 3 z = 7$\par
%~\par
  
	24)\\
  $2 x - 3 y + 5 z = 1$\\
  $3 x + 2 y - z = 4$\\
  $4 x + 7 y - 7 z = 7$\par
~\par
  27)\\
  $m + 6 n + 3 p = 8$\\
  $3 m + 4 n = - 3$\\
  $5 m + 7 n = 1$\par
~\par
  30)\\
  $w - 2 x + 3 y - z = 8$\\
	$w - x - y + z = 4$\\
	$w + x + y + z = 22$\\
	$w - x + y + z = 14$\par
~\par
~\par
~\par
~\par
~\par
~\par
	
\end{multicols}

\pagebreak

\subsection{Applications}\par

	\subsubsection{Value Problems}\par

{\tmstrong{Solve each value problem.}}

1) A collection of dimes and quarters is worth $\$15.25$. There are
103 coins in all.  How many of each is there?\par
~\par
2) A collection of half dollars and nickels is worth $\$13.40$. There
are 34 coins in  all. How many are there?\par
~\par

3) The attendance at a school concert was 578. Admission was $\$2.00$
for adults  and $\$1.50$ for children. The total
receipts were $\$985.00$. How many adults and  how many
children attended?\par
~\par

4) A purse contains $\$3.90$ made up of dimes and quarters. If there
are 21 coins in  all, how many dimes and how many quarters were there?\par
~\par

5) A boy has $\$2.25$ in nickels and dimes. If there are twice as
many dimes as nickels, how many of each kind has he?\par
~\par

6) $\$3.75$ is made up of quarters and half dollars. If the number of
quarters exceeds the number of half dollars by 3,
how many coins of each denomination are there?\par
~\par

7) A collection of 27 coins consisting of nickels and dimes amounts to
$\$2.25$. How many coins of each kind are there?\par
~\par

8) $\$3.25$ in dimes and nickels, were distributed among 45 boys. If
each received one coin, how many received dimes and how many
received nickels?\par
~\par

9) There were 429 people at a play. Admission was $\$1$ each for
adults and 75-cents each for children. The receipts were
$\$372.50$. How many children and how many adults
attended?\par
~\par

10) There were 200 tickets sold for a women's basketball game. Tickets for students were 50-cents each and for adults 75-cents
each. The total amount of  money collected was $\$132.50$.
How many of each type of ticket was sold?\par
~\par

11) There were 203 tickets sold for a volleyball game. For activity-card
holders, the price was $\$1.25$ each and for
noncard holders the price was $\$2$ each. The total
amount of money collected was $\$310$. How many of each type of
ticket was sold?\par
~\par

12) At a local ball game the hotdogs sold for $\$2.50$ each and the
hamburgers sold for $\$2.75$ each. There were 131 total
sandwiches sold for a total value of $\$342$.  How many of
each sandwich was sold?\par
~\par

13) At a recent Vikings game $\$445$ in admission tickets was taken
in. The cost of a student ticket was $\$1.50$ and the
cost of a non-student ticket was $\$2.50$. A total of
232 tickets were sold. How many students and how many non-students attended the game?\par
~\par

14) A bank contains 27 coins in dimes and quarters. The coins have a total
value  of $\$4.95$. Find the number of dimes and quarters
in the bank.\par
~\par

15) A coin purse contains 18 coins in nickels and dimes. The coins have a
total value of $\$1.15$. Find the number of nickels
and dimes in the coin purse.\par
~\par

16) A business executive bought 40 stamps for $\$9.60$. The purchase
included 25-cents stamps and 20-cents stamps. How
many of each type of stamp were bought?\par
~\par

17) A postal clerk sold some 15-cents stamps and some 25-cents
stamps. Altogether, 15 stamps were sold for a total cost
of $\$3.15$. How many of each type of stamps were sold?\par
~\par

18) A drawer contains 15-cents stamps and 18-cents stamps. The
number of 15-cents stamps is  four less than three times the
number of 18-cents stamps. The total value of all the
stamps is $\$1.29$. How many 15-cents stamps are in the
drawer?\par
~\par

19) The total value of dimes and quarters in a bank is $\$6.05$.
There are six more quarters than dimes. Find the number of
each type of coin in the bank.\par
~\par

20) A child's piggy bank contains 44 coins in quarters and dimes. The coins
have a total value of $\$8.60$. Find the number of
quarters in the bank.\par
~\par

21) A coin bank contains nickels and dimes. The number of dimes is 10 less
than twice the number of nickels. The total value of all the coins
is $\$2.75$. Find the number of each type of coin in the
bank.\par
~\par

22) A total of 26 bills are in a cash box. Some of the bills are one dollar
bills, and the rest are five dollar bills. The total amount of cash
in the box is $\$50$. Find  the number of each type of
bill in the cash box.\par
~\par

23) A bank teller cashed a check for $\$200$ using twenty dollar
bills and ten dollar  bills. In all, twelve bills were handed to
the customer. Find the number of twenty dollar bills and
the number of ten dollar bills.\par
~\par

24) A collection of stamps consists of 22-cents stamps and
40-cents stamps. The number of 22-cents stamps is
three more than four times the number of 40-cents stamps. The total value of the stamps is $\$8.34$. Find the number of
22-cents stamps in the collection.\par
~\par

25) A total of $\$27000$ is invested, part of it at 12\% and the rest
at 13\%. The total interest after one year is
$\$3385$. How much was invested at each rate?\par
~\par

26) A total of $\$50000$ is invested, part of it at 5\% and the rest
at 7.5\%. The total interest after one year is $\$3250$.
How much was invested at each rate?\par
~\par

27) A total of $\$9000$ is invested, part of it at 10\% and the rest
at 12\%. The total interest after one year is $\$1030$.
How much was invested at each rate?\par
~\par

28) A total of $\$18000$ is invested, part of it at 6\% and the rest
at 9\%. The total interest after one year is
$\$1248$. How much was invested at each rate?\par
~\par

29) An inheritance of $\$10000$ is invested in 2 ways, part at 9.5\%
and the remainder at 11\%. The combined
annual interest was $\$1038.50$. How much was
invested at each rate?\par
~\par

30) Kerry earned a total of $\$900$ last year on his investments. If
$\$7000$ was invested at a certain rate
of return and $\$9000$ was invested in a fund with a rate that was 2\% higher, find the two rates of interest.\par
~\par

31) Jason earned $\$256$ interest last year on his investments. If
$\$1600$ was invested at a certain rate of return and
$\$2400$ was invested in a fund with a rate that  was
double the rate of the first fund, find the two rates of interest.\par
~\par

32) Millicent earned $\$435$ last year in interest. If
$\$3000$ was invested at a certain  rate of return and
$\$4500$ was invested in a fund with a rate that was 2\% lower, find the two rates of interest.\par
~\par

33) A total of $\$8500$ is invested, part of it at 6\% and the rest
at 3.5\%. The total interest after one year is $\$385$.
How much was invested at each rate?\par
~\par

34) A total of $\$12000$ was invested, part of it at 9\% and the rest
at 7.5\%. The total interest after one year is
$\$1005$. How much was invested at each rate?\par
~\par

35) A total of $\$15000$ is invested, part of it at 8\% and the rest
at 11\%. The total  interest after one year is $\$1455$.
How much was invested at each rate?\par
~\par

36) A total of $\$17500$ is invested, part of it at 7.25\% and the
rest at 6.5\%. The total interest after one year is
$\$1227.50$. How much was invested at each rate?\par
~\par

37) A total of $\$6000$ is invested, part of it at 4.25\% and the
rest at 5.75\%. The total interest after one year is
$\$300$. How much was invested at each rate?\par
~\par

38) A total of $\$14000$ is invested, part of it at 5.5\% and the
rest at 9\%. The total interest after one year is $\$910$.
How much was invested at each rate?\par
~\par

39) A total of $\$11000$ is invested, part of it at 6.8\% and the
rest at 8.2\%. The total interest after one year is
$\$797$. How much was invested at each rate?\par
~\par

40) An investment portfolio earned $\$2010$ in interest last year. If
$\$3000$ was invested at a certain rate of
return and $\$24000$ was invested in a fund with a rate
that was 4\% lower, find the two rates of interest.\par
~\par

41) Samantha earned $\$1480$ in interest last year on her
investments. If $\$5000$ was invested at a certain rate of
return and $\$11000$ was invested in a fund with a rate
that was two-thirds the rate of the first fund, find the two rates of interest.\par
~\par

42) A man has $\$5.10$ in nickels, dimes, and quarters. There are
twice as many nickels as dimes and 3 more dimes than
quarters. How many coins of each kind were there?\par
~\par

43) 30 coins having a value of $\$3.30$ consists of nickels, dimes
and quarters. If there are twice as many quarters as
dimes, how many coins of each kind were there?\par
~\par

44) A bag contains nickels, dimes and quarters having a value of
$\$3.75$. If there are 40 coins in all and 3 times as
many dimes as quarters, how many coins of each kind were there?\par
~\par

%\pagebreak
\newpage

\subsubsection{Mixture Problems}\par

{\tmstrong{Solve each mixture problem.}}

1) A tank contains 8000 liters of a solution that is 40\% acid. How much water should be added to make a solution that is 30\% acid?\par
~\par

2) How much antifreeze should be added to 5 quarts of a 30\% mixture of antifreeze to make a solution that is 50\%
antifreeze?\par
~\par

3) Of 12 pounds of salt water 10\% is salt; of another mixture 3\% is salt.
How many pounds of the second should be added to the first
in order to get a mixture of 5\% salt?\par
~\par

4) How much alcohol must be added to 24 gallons of a 14\% solution of alcohol
in order to produce a 20\% solution?\par
~\par

5) How many pounds of a 4\% solution of borax must be added to 24 pounds of a 12\% solution of borax to obtain a 10\% solution of borax?\par
~\par

6) How many grams of pure acid must be added to 40 grams of a 20\% acid solution to make a solution which is 36\% acid?\par
~\par

7) A 100 LB bag of animal feed is 40\% oats. How many pounds of oats must be added to this feed to produce a mixture which is 50\% oats?\par
~\par

8) A 20 oz alloy of platinum that costs $\$220$ per ounce is mixed
with an alloy that costs $\$400$ per ounce. How many ounces of the $\$400$ alloy should be used to make an
alloy that costs $\$300$ per ounce?\par
~\par

9) How many pounds of tea that cost $\$4.20$ per pound must be mixed
with 12 lb of tea that cost $\$2.25$ per pound to make a
mixture that costs $\$3.40$ per pound?\par
~\par

10) How many liters of a solvent that costs $\$80$ per liter must be
mixed with 6 L of a solvent that costs $\$25$ per liter
to make a solvent that costs $\$36$ per liter?\par
~\par

11) How many kilograms of hard candy that cost $\$7.50$ per kilogram
must be mixed with 24 kg of jelly beans that cost
$\$3.25$ per kilogram to make a mixture
that sells for $\$4.50$ per kilogram?\par
~\par

12) How many kilograms of soil supplement that costs $\$7.00$ per
kilogram must be mixed with 20 kg of aluminum nitrate that
costs $\$3.50$ per kilogram to make a fertilizer
that costs $\$4.50$ per kilogram?\par
~\par

13) How many pounds of lima beans that cost 90-cents per pound must be
mixed with  16 lb of corn that cost 50-cents per pound to make
a mixture of vegetables that costs 65-cents per
pound?\par
~\par

14) How many liters of a blue dye that costs $\$1.60$ per liter must
be mixed with 18 L of a dye that costs $\$2.50$ per
liter to make a mixture that costs $\$1.90$ per liter?\par
~\par

15) Solution A is 50\% acid and solution B is 80\% acid. How much of each
should be used to make 100cc. of a solution that is 68\% acid?\par
~\par

16) A certain grade of milk contains 10\% butter fat and a certain grade of
cream 60\% butter fat. How many quarts of each must be taken so
as to obtain a mixture of 100 quarts that will be 45\%
butter fat?\par
~\par

17) A farmer has some cream which is 21\% butterfat and some which is 15\% butterfat. How many gallons of each must be mixed to
produce 60 gallons of cream which is 19\% butterfat?\par
~\par

18) A syrup manufacturer has some pure maple syrup and some which is 85\% maple syrup. How many liters of each should be mixed to make
150L which  is 96\% maple syrup?\par
~\par

19) A chemist wants to make 50ml of a 16\% acid solution by mixing a 13\% acid solution and an 18\% acid solution. How many milliliters of
each solution should the chemist use?\par
~\par

20) A hair dye is made by blending 7\% hydrogen peroxide solution and a 4\% hydrogen peroxide solution. How many milliliters of each
are used to make a 300 ml solution that is 5\% hydrogen
peroxide?\par
~\par

21) A paint that contains 17\% green dye is mixed with a paint that contains
12\%  green dye. How many gallons of each must be used to make 40
gal of paint that is 15\% green dye?\par
~\par

22) A candy mix sells for $\$2.20$ per kilogram. It contains
chocolates worth $\$1.80$  per kilogram and other candy
worth $\$3.00$ per kilogram. How much of each are in
15 kilograms of the mixture?\par
~\par

23) To make a weed and feed mixture, the Green Thumb Garden Shop mixes fertilizer worth $\$4.00$/lb. with a weed killer
worth $\$8.00$/lb. The mixture will cost
$\$6.00$/lb. How much of each should be used to prepare 500 lb. of the mixture?\par
~\par

24) A grocer is mixing 40 cent per lb. coffee with 60 cent per lb. coffee to
make a mixture worth 54-cents per lb. How much of each kind
of coffee should be used to make 70 lb. of the mixture?\par
~\par

25) A grocer wishes to mix sugar at 9-cents per pound with sugar at 6-cents
per pound to make 60 pounds at 7-cents per pound. What quantity
of each must he take?\par
~\par

26) A high-protein diet supplement that costs $\$6.75$ per pound is
mixed with a vitamin supplement that costs $\$3.25$
per pound. How many pounds of each should be used to make 5
lb of a mixture that costs $\$4.65$ per pound?\par
~\par

27) A goldsmith combined an alloy that costs $\$4.30$ per ounce with
an alloy that costs $\$1.80$per ounce. How many ounces
of each were used to make a mixture of 200 oz costing
$\$2.50$ per ounce?\par
~\par

28) A grocery store offers a cheese and fruit sampler that combines cheddar
cheese that costs $\$8$ per kilogram with kiwis that cost
$\$3$ per kilogram. How many kilograms of each were
used to make a 5 kg mixture that costs $\$4.50$ per kilogram?\par
~\par

29) The manager of a garden shop mixes grass seed that is 60\% rye grass with
70  lb of grass seed that is 80\% rye grass to make a mixture
that is 74\% rye grass. How much of the 60\% mixture
is used?\par
~\par

30) How many ounces of water evaporated from 50 oz of a 12\% salt solution to produce a 15\% salt solution?\par
~\par

31) A caterer made an ice cream punch by combining fruit juice that cost
$\$2.25$ per gallon with ice cream that costs
$\$3.25$ per gallon. How many gallons of each were
used to make 100 gal of punch costing $\$2.50$ per pound?\par
~\par

32) A clothing manufacturer has some pure silk thread and some thread that is 85\% silk. How many kilograms of each must be woven
together to make 75 kg of cloth that is 96\% silk?\par
~\par

33) A carpet manufacturer blends two fibers, one 20\% wool and the second 50\% wool. How many pounds of each fiber should be woven together
to produce 600 lb of a fabric that is 28\% wool?\par
~\par

34) How many pounds of coffee that is 40\% java beans must be mixed with 80 lbs of coffee that is 30\% java beans to make a coffee blend that is
32\% java beans?\par
~\par

35) The manager of a specialty food store combined almonds that cost
$\$4.50$ per pound with walnuts that cost
$\$2.50$ per pound. How many pounds of each were
used to make a 100 lb mixture that cost $\$3.24$ per pound?\par
~\par

36) A tea that is 20\% jasmine is blended with a tea that is 15\% jasmine. How many pounds of each tea are used to make 5 lb of tea
that is 18\% jasmine?\par
~\par

37) How many ounces of dried apricots must be added to 18 oz of a snack mix that contains 20\% dried apricots to make a mixture that
is 25\% dried apricots?\par
~\par

38) How many milliliters of pure chocolate must be added to 150 ml of chocolate topping that is 50\% chocolate to make a topping that is 75\%
chocolate?\par
~\par

39) How many ounces of pure bran flakes must be added to 50 oz of cereal that is 40\% bran flakes to produce a mixture that is 50\% bran
flakes?\par
~\par

40) A ground meat mixture is formed by combining meat that costs $\$2.20$ per pound with meat that costs
$\$4.20$ per pound. How many pounds of each were used to make a 50 lb mixture that costs $\$3.00$ per pound?\par
~\par

41) How many gallons of pure water must be added to 50 gallons of pure acid to make a solution that is 40\% acid?\par
~\par

42) A lumber company combined oak wood chips that cost $\$3.10$ per pound with pine wood chips that cost $\$2.50$ per pound.
How many pounds of each were used to make an 80 lb mixture
costing $\$2.65$ per pound?\par
~\par

43) How many ounces of pure water must be added to 50 oz of a 15\% saline solution to make a saline solution that is 10\% salt?

\newpage

\subsection{Matrix Notation}\par

{\tmstrong{Construct an augmented matrix for each of the systems of equations referenced below.  Then row reduce your matrix to its row echelon form and determine if the given system has (1) no solution, (2) infinitely many solutions, or (3) exactly one solution.  If one solution exists, determine the reduced row echelon form for your matrix and use it to find the solution to the given system.}}\pp

1) - 5): Systems (1) through (5) on page \pageref{matrices1}.\pp

6) - 10): Systems (11) through (15) on page \pageref{matrices1}.\pp

11) - 20): Systems (1) through (10) on page \pageref{matrices2}.\pp

21) - 24): Systems (29) through (32) on page \pageref{matrices3}.%\pp
